\textit{Supplementary materials for}: Feature and order manipulations in a free recall task affect memory for current and future lists

### 2023-01-24 paper/main.tex
entered ECW edits

\author{Jeremy R. Manning\textsuperscript{1, *}, Emily
Whitaker\textsuperscript{1}, Paxton C. Fitzpatrick\textsuperscript{1},
\\Madeline R. Lee\textsuperscript{1}, Allison M. Frantz\textsuperscript{1},
Bryan J. Bollinger\textsuperscript{1},\\Darya Romanova\textsuperscript{1},
Campbell E. Field\textsuperscript{1}, and Andrew C. Heusser\textsuperscript{1,
2}\\\textsuperscript{1}Dartmouth College\\\textsuperscript{2}Akili
Interactive\\\textsuperscript{*}Corresponding author:
jeremy.r.manning@dartmouth.edu}

@@

Our understanding of how we form situation models and schemas, and how they
interact with our subjective experiences and memories, is constrained in part
by substantial differences in how we study these processes. Situation models
and schemas are most often studied using ``naturalistic'' stimuli such as
narratives and movies~\citep{ZwaaEtal95,ZwaaRadv98, NastEtal20}. In contrast,
our understanding of how we organize our memories has been most widely informed
by more traditional paradigms like free recall of random word
lists~\citep{Kaha12, Kaha20}. In free recall, participants study lists of items
and are instructed to recall the items in any order they choose. The orders in
which words come to mind can provide insights into how participants have
organized their memories of the studied words. Because random word lists are
unstructured by design, it is not clear if or how non-trivial situation models
might apply to these stimuli. Nevertheless, there are \textit{some}
commonalities between memory for word lists and memory for real-world
experiences.

@@

Over the past half-century, context-based models have enjoyed impressive
success at explaining many stereotyped behaviors observed during free recall
and other list-learning tasks~\citep{Este55a, RaaiShif80, GlenEtal83,
HowaKaha02a, SiroEtal05, KimbEtal07, PolyKaha08, SedeEtal08, PolyEtal09,
ShanHowa12}. These phenomena include the well-known recency and primacy effects
(superior recall of items from the end and, to a lesser extent, from the
beginning of the study list), as well as semantic and temporal clustering
effects~\citep{KahaEtal08b}. The contiguity effect is an example of temporal
clustering, which is perhaps the dominant form of organization in free recall.
This effect can be seen in the tendency for people to successively recall items
that occupied neighboring positions in the studied list~\citep{Kaha96}. There
are also striking effects of semantic clustering~\citep{RomnEtal93, Bous53,
BousEtal54, JenkRuss52, MannKaha12}, whereby the recall of a given item is more
likely to be followed by recall of a similar or related item than a dissimilar
or unrelated one. In general, people organize memories for words along a wide
variety of stimulus dimensions. As formalized by models like the
\textit{Context Maintenance and Retrieval Model}~\citep{PolyEtal09}, the
stimulus features associated with each word (e.g.\ the word's meaning, font
size, font color, location on the screen, size of the object the word
represents, etc.) are incorporated into the participant's mental context
representation~\citep{SmitVela01, MannEtal15, Mann20, MannEtal11, MannEtal12}.
During a memory test, any of these features may serve as a memory cue, which in
turn leads the participant to recall in succession words that share stimulus
features.

@@

Participants received course credit for enrolling in our study. We asked each
participant to fill out a demographic survey that included questions about
their age, gender, ethnicity, race, education, vision, reading
impairments, medications or recent injuries, coffee consumption on the day of
testing, and level of alertness at the time of testing. All components of the
demographics survey were optional. One participant elected not to fill out any
part of the demographic survey, and all other participants answered some or all
of the survey questions.

@@

\paragraph{Constructing feature-sorted lists.} Given a list of words, a
relevant feature, and each word's value(s) for that feature, we developed a
stochastic algorithm for (noisily) sorting the words. The stochastic aspect of
our sorting procedure enabled us to obtain unique orderings for each
participant. First, we choose a word uniformly and at random from the set of
candidates. Next, we compute the distances between the chosen word's feature(s)
and the corresponding feature(s) of all yet-to-be-presented words. Third, we
convert these distances (between the previously presented word's feature
values, $a$, and the candidate word's feature values, $b$) to similarity
scores:

@@

Suppose that two lists contain unequal numbers of items of each size. For
example, suppose that list $A$ contains all ``large'' items, whereas list $B$
contains an equal mix of ``large'' and ``small'' items. For a participant
recalling list $A$, any correctly recalled item will necessarily match the size
of the previous correctly recalled item. In other words, successively recalling
several list $A$ items of the same size is essentially meaningless, since
\textit{any} correctly recalled list $A$ word will be large. In contrast,
successively recalling several list $B$ items of the same size \textit{could}
be meaningful, since (early in the recall sequence) the yet-to-be-recalled
items come from a mix of sizes. However, once all of the small items on list
$B$ have been recalled, the best possible next matching recall will be a large
item. And all subsequent correct recalls must also be large items-- so for
those later recalls it becomes difficult to determine whether the participant
is successively recalling large items because they are organizing their
memories according to size, or (alternatively), whether they are simply
recalling the yet-to-be-recalled items in a random order. In general, the
precise order and blend of feature values expressed in a given list, the orders
and numbers of correct recalls a participant makes, the number of intervening
presentation positions between successive recalls, and so on, can all affect
the range of clustering scores that are possible to observe for a given list.
An uncorrected clustering score therefore conflates participants' actual memory
organization with other ``nuisance'' factors.

@@

We found that \textit{adding} irrelevant visual features on later lists that
had not been present on early lists (as in the reduced (early) condition)
served to enhance recall performance relative to conditions where all lists had
the same blends of features (accuracy for feature rich versus reduced (early):
$t(107) = -2.230, p = 0.028$; reduced versus reduced (early): $t(101) = -2.045,
p = 0.043$; also see Fig.~\accuracyByList A). However, \textit{subtracting}
irrelevant visual features on later lists that \textit{had} been present on
early lists (as in the reduced (late) condition) did not appear to impact
recall performance (accuracy for feature rich versus reduced (late): $t(106) =
-0.638, p = 0.525$; reduced versus reduced (late): $t(100) = -0.407, p =
0.685$). These comparisons suggest that recall accuracy has a directional
component (i.e., accuracy is affected differently by removing features later
that had been present earlier versus adding features later that had
\textit{not} been present earlier). In contrast, we found that participants
exhibited more temporal and feature-based clustering when we added irrelevant
visual features to \textit{any} lists (comparisons of clustering on feature
rich anversusd reduced lists are reported above; temporal clustering in reduced
versus reduced (early) and reduced versus reduced (late) conditions: $t$s $\leq
-9.780$, $p$s $< 0.001$; feature-based clustering in reduced versus reduced
(early) and reduced versus reduced (late) conditions: $t$s $\leq -5.443$, $p$s
$< 0.001$). Temporal and feature-based clustering were not reliably different
in the feature rich, reduced (early), and reduced (late) conditions (temporal
clustering in feature rich versus reduced (early) and feature rich versus
reduced (late) conditions: $t$s $\geq -1.434$, $p$s $\geq 0.154$; feature-based
clustering in feature rich versus reduced (early) and feature rich versus
reduced (late) conditions: $t$s $\geq -1.359$, $p$s $> 0.177$).

@@

The above comparisons between memory performance on early lists in the order
manipulation versus feature rich conditions highlight how sorted lists are
remembered differently from random lists. We also wondered how sorting lists
along each feature dimension influenced memory relative to sorting lists along
the other feature dimensions. Participants trended towards remembering early
lists that were sorted semantically better than lexicographically sorted lists
($t(118) = 1.936, p = 0.055$). Participants also remembered visually sorted
lists better than lexicographically sorted lists ($t(119) = 2.145, p = 0.034$).
However, participants showed no reliable differences in recall for semantically
versus visually sorted lists ($t(119) = 0.113, p = 0.910$). Participants
temporally clustered semantically sorted lists more strongly than either
lexicographically ($t(118) = 5.572, p < 0.001$) or visually ($t(119) = 6.215, p
< 0.001$) sorted lists, but did not show reliable differences in temporal
clustering on lexicographically versus visually sorted lists ($t(119) = 0.189,
p = 0.850$). Participants also showed reliably more semantic clustering on
semantically sorted lists than lexicographically (category: $t(118) = 3.492, p
= 0.001$, size: $t(118) = 3.972, p < 0.001$) or visually (category: $t(119) =
2.702, p = 0.008$, size: $t(119) = 4.230, p < 0.001$) sorted lists; more
lexicographic clustering on lexicographically sorted lists than semantically
(length: $t(118) = 3.112, p = 0.002$; first letter: $t(118) = 3.686, p =
0.000$) or visually (length: $t(119) = 3.024, p = 0.003$; first letter: $t(119)
= 2.644, p = 0.009$) sorted lists; and more visual clustering on visually
sorted lists than semantically (color: $t(119) = -2.659, p = 0.009$; location:
$t(119) = -4.604, p < 0.001$) or lexicographically (color: $t(119) = -2.366, p
= 0.020$; location: $t(119) = -4.265, p < 0.001$) sorted lists. In summary,
sorting lists by different features appeared to have slightly different effects
on overall memory performance and temporal clustering, and people tended to
cluster their recalls along a given feature dimension more when the studied
lists were (versus were not) sorted along that dimension.

@@

Conceptualization: JRM and ACH. Methodology: JRM and ACH. Software: JRM, PCF,
CEF, and ACH. Analysis: JRM, PCF, and ACH. Data collection: EW, PCF, MRL, AMF,
BJB, DR, and CEF. Data curation and management: EW, PCF, MRL, ACH. Writing
(original draft): JRM. Writing (review and editing): EW, PCF, MRL, AMF, BJB,
DR, CEF, and ACH. Supervision: JRM and ACH. Project administration: EW, and
PCF. Funding acquisition: JRM.

>>>>

\author{Jeremy R. Manning\textsuperscript{1, *}, Emily C.
Whitaker\textsuperscript{1}, Paxton C. Fitzpatrick\textsuperscript{1},
\\Madeline R. Lee\textsuperscript{1}, Allison M. Frantz\textsuperscript{1},
Bryan J. Bollinger\textsuperscript{1},\\Darya Romanova\textsuperscript{1},
Campbell E. Field\textsuperscript{1}, and Andrew C. Heusser\textsuperscript{1,
2}\\\textsuperscript{1}Dartmouth College\\\textsuperscript{2}Akili
Interactive\\\textsuperscript{*}Corresponding author:
jeremy.r.manning@dartmouth.edu}

@@

Our understanding of how we form situation models and schemas, and how they
interact with our subjective experiences and memories, is constrained in part
by substantial differences in how we study these processes. Situation models
and schemas are most often studied using ``naturalistic'' stimuli such as
narratives and movies~\citep{ZwaaEtal95,ZwaaRadv98, NastEtal20}. In contrast,
our understanding of how we organize our memories has been most widely informed
by more traditional paradigms like free recall of random word
lists~\citep{Kaha12, Kaha20}. In free recall, participants study lists of items
and are instructed to recall the items in any order they choose. The orders in
which words come to mind can provide insights into how participants have
organized their memories of the studied words. Because random word lists are
unstructured by design, it is not clear if, or how, non-trivial situation models
might apply to these stimuli. Nevertheless, there are \textit{some}
commonalities between memory for word lists and memory for real-world
experiences.

@@

Over the past half-century, context-based models have had impressive success at
explaining many stereotyped behaviors observed during free recall and other
list-learning tasks~\citep{Este55a, RaaiShif80, GlenEtal83, HowaKaha02a,
SiroEtal05, KimbEtal07, PolyKaha08, SedeEtal08, PolyEtal09, ShanHowa12}. These
phenomena include the well-known recency and primacy effects (superior recall
of items from the end and, to a lesser extent, from the beginning of the study
list), as well as semantic and temporal clustering effects~\citep{KahaEtal08b}.
The contiguity effect is an example of temporal clustering, which is perhaps
the dominant form of organization in free recall. This effect can be seen in
the tendency for people to successively recall items that occupied neighboring
positions in the studied list~\citep{Kaha96}. There are also striking effects
of semantic clustering~\citep{RomnEtal93, Bous53, BousEtal54, JenkRuss52,
MannKaha12}, whereby the recall of a given item is more likely to be followed
by recall of a similar or related item than a dissimilar or unrelated one. In
general, people organize memories for words along a wide variety of stimulus
dimensions. As formalized by models like the \textit{Context Maintenance and
Retrieval Model}~\citep{PolyEtal09}, the stimulus features associated with each
word (e.g.\ the word's meaning, size of the object the word represents, the
letters that make up the word, font size, font color, location on the screen,
etc.) are incorporated into the participant's mental context
representation~\citep{SmitVela01, MannEtal15, Mann20, MannEtal11, MannEtal12}.
During a memory test, any of these features may serve as a memory cue, which in
turn leads the participant to recall in succession words that share stimulus
features.

@@

Participants either received course credit or a one-time \$10 payment for
enrolling in our study. We asked each participant to fill out a demographic
survey that included questions about their age, gender, ethnicity, race,
education, vision, reading impairments, medications or recent injuries, coffee
consumption on the day of testing, and level of alertness at the time of
testing. All components of the demographics survey were optional. One
participant elected not to fill out any part of the demographic survey, and all
other participants answered some or all of the survey questions.

@@

\paragraph{Constructing feature-sorted lists.} Given a list of words, a
relevant feature, and each word's value(s) for that feature, we developed a
stochastic algorithm for (noisily) sorting the words. The stochastic aspect of
our sorting procedure enabled us to obtain unique orderings for each
participant. First, we choose a word uniformly and at random from the set of
candidates. Second, we compute the distances between the chosen word's feature(s)
and the corresponding feature(s) of all yet-to-be-presented words. Third, we
convert these distances (between the previously presented word's feature
values, $a$, and the candidate word's feature values, $b$) to similarity
scores:

@@

Suppose that two lists contain unequal numbers of items of each size. For
example, suppose that list $A$ contains all ``large'' items, whereas list $B$
contains an equal mix of ``large'' and ``small'' items. For a participant
recalling list $A$, any correctly recalled item will necessarily match the size
of the previous correctly recalled item. In other words, successively recalling
several list $A$ items of the same size is essentially meaningless, since
\textit{any} correctly recalled list $A$ word will be large. In contrast,
successively recalling several list $B$ items of the same size \textit{could}
be meaningful, since (early in the recall sequence) the yet-to-be-recalled
items come from a mix of sizes. However, once all of the small items on list
$B$ have been recalled, the best possible next matching recall will be a large
item. All subsequent correct recalls must also be large items-- so for
those later recalls it becomes difficult to determine whether the participant
is successively recalling large items because they are organizing their
memories according to size, or (alternatively), whether they are simply
recalling the yet-to-be-recalled items in a random order. In general, the
precise order and blend of feature values expressed in a given list, the order
and number of correct recalls a participant makes, the number of intervening
presentation positions between successive recalls, and so on, can all affect
the range of clustering scores that are possible to observe for a given list.
An uncorrected clustering score therefore conflates participants' actual memory
organization with other ``nuisance'' factors.

@@

We found that \textit{adding} irrelevant visual features on later lists that
had not been present on early lists (as in the reduced (early) condition)
served to enhance recall performance relative to conditions where all lists had
the same blends of features (accuracy for feature rich versus reduced (early):
$t(107) = -2.230, p = 0.028$; reduced versus reduced (early): $t(101) = -2.045,
p = 0.043$; also see Fig.~\accuracyByList A). However, \textit{subtracting}
irrelevant visual features on later lists that \textit{had} been present on
early lists (as in the reduced (late) condition) did not appear to impact
recall performance (accuracy for feature rich versus reduced (late): $t(106) =
-0.638, p = 0.525$; reduced versus reduced (late): $t(100) = -0.407, p =
0.685$). These comparisons suggest that recall accuracy has a directional
component: accuracy is affected differently by removing features later
that had been present earlier versus adding features later that had
\textit{not} been present earlier. In contrast, we found that participants
exhibited more temporal and feature-based clustering when we added irrelevant
visual features to \textit{any} lists (comparisons of clustering on feature
rich anversusd reduced lists are reported above; temporal clustering in reduced
versus reduced (early) and reduced versus reduced (late) conditions: $t$s $\leq
-9.780$, $p$s $< 0.001$; feature-based clustering in reduced versus reduced
(early) and reduced versus reduced (late) conditions: $t$s $\leq -5.443$, $p$s
$< 0.001$). Temporal and feature-based clustering were not reliably different
in the feature rich, reduced (early), and reduced (late) conditions (temporal
clustering in feature rich versus reduced (early) and feature rich versus
reduced (late) conditions: $t$s $\geq -1.434$, $p$s $\geq 0.154$; feature-based
clustering in feature rich versus reduced (early) and feature rich versus
reduced (late) conditions: $t$s $\geq -1.359$, $p$s $> 0.177$).

@@

The above comparisons between memory performance on early lists in the order
manipulation versus feature rich conditions highlight how sorted lists are
remembered differently from random lists. We also wondered how sorting lists
along each feature dimension influenced memory relative to sorting lists along
the other feature dimensions. Participants trended towards remembering early
lists that were sorted semantically better than lexicographically sorted lists
($t(118) = 1.936, p = 0.055$). Participants also remembered visually sorted
lists better than lexicographically sorted lists ($t(119) = 2.145, p = 0.034$).
However, participants showed no reliable differences in recall for semantically
versus visually sorted lists ($t(119) = 0.113, p = 0.910$). Participants
temporally clustered semantically sorted lists more strongly than either
lexicographically ($t(118) = 5.572, p < 0.001$) or visually ($t(119) = 6.215, p
< 0.001$) sorted lists, but did not show reliable differences in temporal
clustering on lexicographically versus visually sorted lists ($t(119) = 0.189,
p = 0.850$). Participants also showed reliably more semantic clustering on
semantically sorted lists than lexicographically (category: $t(118) = 3.492, p
= 0.001$, size: $t(118) = 3.972, p < 0.001$) or visually (category: $t(119) =
2.702, p = 0.008$, size: $t(119) = 4.230, p < 0.001$) sorted lists; more
lexicographic clustering on lexicographically sorted lists than semantically
(length: $t(118) = 3.112, p = 0.002$; first letter: $t(118) = 3.686, p =
0.000$) or visually (length: $t(119) = 3.024, p = 0.003$; first letter: $t(119)
= 2.644, p = 0.009$) sorted lists; and more visual clustering on visually
sorted lists than semantically (color: $t(119) = -2.659, p = 0.009$; location:
$t(119) = -4.604, p < 0.001$) or lexicographically (color: $t(119) = -2.366, p
= 0.020$; location: $t(119) = -4.265, p < 0.001$) sorted lists. In summary,
sorting lists by different features appeared to have slightly different effects
on overall memory performance and temporal clustering.  Participants also tended to
cluster their recalls along a given feature dimension more when the studied
lists were (versus were not) sorted along that dimension.

@@

Conceptualization: JRM and ACH. Methodology: JRM and ACH. Software: JRM, PCF,
CEF, and ACH. Analysis: JRM, PCF, and ACH. Data collection: ECW, PCF, MRL, AMF,
BJB, DR, and CEF. Data curation and management: ECW, PCF, MRL, and ACH. Writing
(original draft): JRM. Writing (review and editing): ECW, PCF, MRL, AMF, BJB,
DR, CEF, and ACH. Supervision: JRM and ACH. Project administration: ECW and
PCF. Funding acquisition: JRM.

### 2023-01-24 paper/supplement.tex
entered ECW edits

    \caption{\textbf{Recall accuracy by study list number.}  Each panel displays the average recall accuracy (across participants) as a function of the number of studied lists for the
    random conditions (\textbf{A.}), order manipulation conditions (\textbf{B.}), and adaptive conditions (\textbf{C.}.  The conditions are denoted by color.  Note that the first four
    lists in all of the ``adaptive'' conditions were ordered randomly to compute a baseline fingerprint for each participant prior to initiating the adaptive ordering procedure.  All error
    ribbons denote bootstrap-estimated 95\% confidence intervals, calculated across participants.}
        \label{fig:accuracy-by-list}
\end{figure}

>>>>

    \caption{\textbf{Recall accuracy by study list number.}  Each panel displays the average recall accuracy (across participants) as a function of the number of studied lists for the
    random conditions (\textbf{A.}), order manipulation conditions (\textbf{B.}), and adaptive conditions (\textbf{C.}).  The conditions are denoted by color.  Note that the first four
    lists in all of the ``adaptive'' conditions were ordered randomly to compute a baseline fingerprint for each participant prior to initiating the adaptive ordering procedure.  All error
    ribbons denote bootstrap-estimated 95\% confidence intervals, calculated across participants.}
        \label{fig:accuracy-by-list}
\end{figure}
